\bf On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2024-11-23 Manuscript/manuscript.tex
comments on gg3b results
went over referee 2 comments

Based on an exploratory analysis, the reaction times are adequately
modeled by a Generalized Gamma (GG) distribution with random effects in
model parameters. The GG distribution has three parameters, denoted by
$\text{GG}(\mu, \sigma, \nu)$ has density function
\[
f_Y(y \mid \mu, \sigma, \nu) = 
\frac{|\nu| \theta^\theta z^{\theta}}{\Gamma(\theta) y} 
\exp\left(-z \theta\right),
\]
for $y > 0$, $\mu > 0$, $\sigma > 0$, and $\nu \neq 0$,
where $z = \left(\frac{y}{\mu}\right)^\nu$,
$\theta = 1 / (\sigma^2 \nu^2)$, and
$\Gamma(\cdot)$ denotes the Gamma function.
The GG distribution is highly flexible, encompassing several  
well-known distributions as special cases, such as the
Weibull ($\mu = \nu)$ and  Gamma $(\nu = 1)$ distributions.
Its mean is
\[
  \mu \frac{\Gamma(\theta + 1 / \nu)}
  {\theta^{1 / \nu} \Gamma(\theta)},
\]
provided $\theta > -1 / \nu$. Here,
$\mu$ scales the central tendency, $\sigma$ controls 
dispersion, and $\nu$ determines skewness. This parameterization allows 
the distribution to model asymmetric and heavy-tailed data effectively, making 
it particularly suitable for reaction times.
An implementation of this distribution is available from R package
\texttt{gamlss.dist} \citep{rigby2019distributions}.

@@

\jy{Owen: an additional figure of the GG density without random
  effects (use \texttt{curve(dGG(x, mu = , sigma = , nu =  ), 0, 0.03)}) and
  with random effects (through \texttt{density()})}
\of{Please see figure 5}

@@

@@

\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than threshold 0.08,
  0.09, and 0.10 seconds based on the two effect model.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & Threshold 0.08 & Threshold 0.09 & Threshold 0.10  \\ 
   \midrule
   Excluding 2022 & $5.31\cdot10^{-5}$ & $3.53\cdot10^{-4}$ &  $1.94\cdot10^{-3}$  \\ 
   Including 2022 & $6.84\cdot10^{-5}$ & $4.95\cdot10^{-4}$ & $2.76\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}
\of{table has been updated with gg3b results}

>>>>

Based on an exploratory analysis, the reaction times are adequately
modeled by a Generalized Gamma (GG) distribution with random effects in
model parameters. The GG distribution has three parameters, denoted by
$\text{GG}(\mu, \sigma, \nu)$ has density function
\begin{equation}
  \label{eq:gg}
f_Y(y \mid \mu, \sigma, \nu) = 
\frac{|\nu| \theta^\theta z^{\theta}}{\Gamma(\theta) y} 
\exp\left(-z \theta\right),
\end{equation}
for $y > 0$, $\mu > 0$, $\sigma > 0$, and $\nu \neq 0$,
where $z = \left(\frac{y}{\mu}\right)^\nu$,
$\theta = 1 / (\sigma^2 \nu^2)$, and
$\Gamma(\cdot)$ denotes the Gamma function.
The GG distribution is highly flexible, encompassing several  
well-known distributions as special cases, such as the
Weibull ($\mu = \nu)$ and  Gamma $(\nu = 1)$ distributions.
Its mean is
\[
  \mu \frac{\Gamma(\theta + 1 / \nu)}
  {\theta^{1 / \nu} \Gamma(\theta)},
\]
provided $\theta > -1 / \nu$. Here,
$\mu$ scales the central tendency, $\sigma$ controls 
dispersion, and $\nu$ determines skewness. This parameterization allows 
the distribution to model asymmetric and heavy-tailed data effectively, making 
it particularly suitable for reaction times.
An implementation of this distribution is available from R package
\texttt{gamlss.dist} \citep{rigby2019distributions}.

@@

\jy{Owen: an additional figure of the GG density without random
  effects (use \texttt{curve(dGG(x, mu = , sigma = , nu =  ), 0, 0.03)}) and
  with random effects (through \texttt{density()})}
\of{Please see figure 5}
\jy{This cannot be right. Black is always above blue. Both need to
  integrate to one.}

@@

\jy{Use aspect ratio 1 for QQ plot}

@@

\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than threshold 0.08,
  0.09, and 0.10 seconds based on the two effect model.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & Threshold 0.08 & Threshold 0.09 & Threshold 0.10  \\ 
   \midrule
   Excluding 2022 & $5.31\cdot10^{-5}$ & $3.53\cdot10^{-4}$ &  $1.94\cdot10^{-3}$  \\ 
   Including 2022 & $6.84\cdot10^{-5}$ & $4.95\cdot10^{-4}$ & $2.76\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}
\of{table has been updated with gg3b results}
\jy{Could you also test sensitivity for excluding the positive but
  fault reaction times?}

### 2024-11-23 Manuscript/reply.tex
went over referee 2 comments

We have revised the Introduction and Discussion sections to clearly
outline the goals of the study. Both reviewers noted issues with
organization, so we streamlined the manuscript to emphasize the two
main objectives: (1) identifying inconsistencies in the 2022 reaction
times and (2) assessing whether the 0.1-second threshold is fair. We
removed unrelated sections to keep the focus on these central
points. Regarding the rank comparison and timing devices at the 2022
Championships, we intentionally used cautious language, as there is no
evidence to claim a miscalibration or fault in Seiko’s timing
devices. To avoid potential legal risks, we noted the unusually fast
reaction times in 2022 without making unsubstantiated claims. Instead,
we allow readers to interpret the underlying cause of the anomaly
based on the presented data.

@@

If you are trying to answer question (1), you should include reaction times from
every spring event, not just the hurdles. If you are trying to answer question
2, why not do the following Monte Carlo test:

1. Sample a single reaction time for each athlete in the final.
2. Calculate the mean reaction time for the simulated final.
3. Compute the probability of seeing the observed average reaction time deviate
as much as it did in the 2022 Championships.

This test seems better than a rank sum test, as ranks do not capture the
differences between times.

\end{comment}

Thank you for the feedback. We are attempting to address question~(1)
and have expanded our analysis to include reaction times from
additional sprint events at the World Championships, including both
men and women's 100-meter data. This broader dataset helps provide a
more comprehensive assessment of whether the venue in 2022 was
unusually fast.

Please see our second response to Reviewer 1 for additional 
comments relating to longer sprint events (200-meter or longer), 
and why these were not also included in our analyses.

@@

Thank you for raising these points. Regarding the exclusion of
disqualified runners, we chose not to include athletes with negative
reaction times, as these indicate an attempt to predict the starting
gun rather than a true reaction. Negative reaction times do not align
with the purpose of the GLMM analysis, which aims to evaluate the
feasibility of the 0.1-second reaction threshold. Including negative
times would introduce results where the reaction occurred before the
gun fired, making them unsuitable for this analysis.

For the approximately 20 disqualified athletes with positive reaction
times, we examined the impact of including them in the analysis. The
probability of observing a reaction time below 0.1 seconds increased
\jy{Are these based on the expanded dataset or the original dataset?}
slightly from 0.0029 to 0.0039, with similar parameter estimates for
the random effects. Given these results, we opted to include these
observations in the dataset.

Distribution: \jy{need to check R package gamlss.}

\jy{Find these references and cite them:
Sprint Start Regulation in Athletics: A Critical Review.
Milloz M, Hayes K, Harrison AJ.
Sports Med. 2021 Jan;51(1):21-31. doi: 10.1007/s40279-020-01350-4.
PMID: 33125639 Review.
Reaction time aspects of elite sprinters in athletic world championships.
Tønnessen E, Haugen T, Shalfawi SA.
J Strength Cond Res. 2013 Apr;27(4):885-92. doi: 10.1519/JSC.0b013e31826520c3.
PMID: 22739331
}

@@

Thank you for this suggestion. Unfortunately, comprehensive data on
Devon Allen's reaction times is not readily available. We gathered as
much data as we could find online, but there is insufficient
information from lower-level meets, as reaction times are typically
only recorded at major competitions (e.g., national and world
championships). Given the limited data, we could not conduct a formal
statistical analysis of his reaction times. However, we have added a
brief discussion of this issue in paragraph~???] of the Discussion Section.

@@

We added additional explanations to address this in the introductions section.
\jy{Give paragraph number for specific locations}

@@

The nomenclature for the greek variables in Table 2 are defined in the GLMM
subsection of the methods section (3.2).
\eds{let's revisit this again after objective 2 analyses are set}

@@

Yes, the comments about the difficulty of data gathering are meant to demonstrate
that we worked hard to find as much data as we could and, as a note to World
Athletics (track and field governing body), that they need to do a better job
at data standardization.  Currently the races on their website do not include
reaction times, which meant that we had to look in alternative places to perform
the analyses that we did.  The data we found from national level meets in 2022
was difficult to come by as it required searching for every country's championship
results individually and when we did find it, it was often in the native language
making it harder to understand and interpret.

@@

We included this plot to show the venue effect for each year and to visually
demonstrate how 2022 compares to other years.  We added additional comments
about the probability of observing an extreme venue effect to provide
quantitative data supporting our argument.
\eds{let's revisit this again after objective 2 analyses are set}

>>>>

Thank you for your valuable feedback. We have thoroughly revised the
Introduction and Discussion sections to clearly articulate the study's
objectives. As both reviewers noted organizational issues, we
streamlined the manuscript to focus on its two main goals: (1)
identifying inconsistencies in reaction times at the 2022 World
Championships, and (2) assessing whether the 0.1-second threshold is
statistically justified. Unrelated sections were removed to maintain a
clear and focused narrative.

Regarding the rank comparison and timing devices at the 2022
Championships, we intentionally used cautious language, as there is no
direct evidence to suggest miscalibration or faults in Seiko’s timing
devices. To avoid potential legal risks, we emphasized the observed
anomaly of unusually fast reaction times in 2022 without making
unsubstantiated claims. Instead, we presented the data transparently,
allowing readers to interpret the possible underlying causes of the
anomaly.

@@

If you are trying to answer question (1), you should include reaction times from
every spring event, not just the hurdles. If you are trying to answer
question~2, why not do the following Monte Carlo test:\\
\indent 1. Sample a single reaction time for each athlete in the final.\\
\indent 2. Calculate the mean reaction time for the simulated final.\\
\indent 3. Compute the probability of seeing the observed average reaction time deviate
as much as it did in the 2022 Championships.\\

This test seems better than a rank sum test, as ranks do not capture the
differences between times.
\end{comment}

Thank you for the feedback. We clarify that our focus is question~(1): 
evaluating whether the venue in 2022 was unusually fast. To address this, 
we expanded our analysis to include reaction times from additional sprint 
events at the World Championships, including both men’s and women’s 
100-meter dashes. This broader dataset provides a more comprehensive 
assessment of venue effects in 2022.

Your suggestion to include reaction times from more sprint events aligns 
with feedback from Reviewer~1.  While the 100-meter dash reaction times 
showed no significant differences from the 110-meter hurdles, we excluded 
200-meter and longer dashes due to their significantly different reaction 
times. These longer events represent distinct contexts unsuitable for this 
analysis.

@@

Thank you for raising these concerns. Regarding the exclusion of 
disqualified runners, we chose not to include athletes with negative 
reaction times, as these represent attempts to predict the starting gun 
rather than true reactions. Negative reaction times do not align with 
our objective of evaluating the feasibility of the 0.1-second threshold, 
as they reflect pre-emptive behavior rather than genuine reaction
times.

The Gamma GLMM model has been replaced with a generalized Gamma 
distribution within the GAMLSS framework, incorporating random effects 
at the venue and heat levels in its distributional parameters. This 
revised model provides improved and adequate fit to the data, as shown 
by diagnostic plots in Figure~3.

For the approximately 20 disqualified athletes with positive reaction
times, we examined the impact of including them in the analysis. The
probability of observing a reaction time below 0.1 seconds increased
\jy{Are these based on the expanded dataset or the original dataset?}
\jy{These numbers need to be updated}
slightly from 0.0029 to 0.0039, with similar parameter estimates for
the random effects. Given these results, we opted to include these
observations in the dataset.

@@

Thank you for this suggestion. Unfortunately, comprehensive data on 
Devon Allen’s reaction times is not readily available. While we gathered 
as much data as possible from publicly available sources, reaction times 
are typically recorded only at major competitions, such as national and 
world championships. As a result, the dataset is too limited to support 
a formal statistical analysis of his reaction times.
To address this point, we have added a brief discussion in 
paragraph~??? of the Discussion Section, noting the limitations in data 
availability and the implications of this issue.

@@

We added additional explanations to address this in the introductions section.
\jy{Owen: Which paragraph?}

@@

Thank you for pointing out the need for clearer intrepretation of
Table~2.
\jy{Owen: please add standard errors of the three intercept  to the table.}
The table now contains fitted parameters of the generalized Gamma
distribution GAMLSS as specified Equations (1)--(3). The table caption has been
expanded to provide detailed explanations of the fitted parameters.

@@

Yes, the comments about the difficulty of data gathering are included
to highlight the effort required to compile a comprehensive
dataset. This also serves as a note to World Athletics, emphasizing
the need for better data standardization. Currently, race results on
their website do not include reaction times, requiring us to search
alternative sources. Gathering data from national-level meets in 2022
involved searching each country’s championship results individually,
often encountering language barriers that complicated interpretation.

@@

Thank you for the feedback. The purpose of Figure~3 is to display the venue 
effects for each year and visually illustrate how 2022 compares to other years. 
To address your concern, we added additional comments quantifying the 
probability of observing an extreme venue effect, which strengthens the 
argument that 2022 is anomalous. If needed, we are open to revising the plot 
further or integrating it into a single panel showing deltas, as suggested.
\eds{let's revisit this again after objective 2 analyses are set}
